---
title: "Exercise 01 – Missing Data Analysis with ACS Data"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../output"
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

In this exercise, we use the file `acs.RData`. It contains two datasets: `acs`
(fully observed) and `acs.mis` (with nonresponse). Both are derived from the
`acs12` dataset included in the R package `openintro`. The data contain a small
subset of variables and observations from the American Community Survey (ACS)
2012.

```{r load-data}
load("../acs.RData")
```

---

# Part (a): Identifying Nonresponse

**Question:** Which variables are affected by nonresponse in `acs.mis`? What is
the nonresponse rate for these variables?

## Answer

```{r part-a}
# Count NAs per variable
na_counts <- colSums(is.na(acs.mis))
na_rates  <- colMeans(is.na(acs.mis))

# Display results
data.frame(
  Variable       = names(na_counts),
  Missing_Count  = na_counts,
  Missing_Rate   = paste0(round(na_rates * 100, 1), "%"),
  row.names      = NULL
)
```

**Discussion:**

Two variables are affected by nonresponse:

- **`income`**: 549 missing values out of 1605 observations (34.2%).
- **`married`**: 504 missing values out of 1605 observations (31.4%).

All other variables (`employment`, `race`, `age`, `gender`, `edu`) are fully
observed.

---

# Part (b): Average Income from Fully Observed Data

**Question:** Compute the average income and its standard error (assuming
simple random sampling with replacement) using the fully observed data `acs`.

## Answer

Under simple random sampling with replacement, the standard error of the sample
mean is:

$$
\text{SE}(\bar{x}) = \frac{s}{\sqrt{n}}
$$

```{r part-b}
mean_income_full <- mean(acs$income)
se_income_full   <- sd(acs$income) / sqrt(nrow(acs))

cat("Full data (acs):\n")
cat("  Mean income :", round(mean_income_full, 2), "\n")
cat("  SE          :", round(se_income_full, 2), "\n")
cat("  n           :", nrow(acs), "\n")
```

---

# Part (c): Average Income from `acs.mis` — CCA vs. ACA

**Question:** Compute the average income and its standard error using
`acs.mis`. Compute these statistics once using complete-case analysis (CCA) and
once using available-case analysis (ACA). Discuss your findings. What do you
conclude regarding the nonresponse mechanism?

## Answer

### Complete-Case Analysis (CCA)

CCA discards all rows that have **any** missing value across all variables used
in the analysis.

```{r part-c-cca}
acs_complete <- acs.mis[complete.cases(acs.mis), ]

mean_income_cca <- mean(acs_complete$income)
se_income_cca   <- sd(acs_complete$income) / sqrt(nrow(acs_complete))

cat("Complete-Case Analysis (CCA):\n")
cat("  Mean income :", round(mean_income_cca, 2), "\n")
cat("  SE          :", round(se_income_cca, 2), "\n")
cat("  n used      :", nrow(acs_complete), "\n")
```

### Available-Case Analysis (ACA)

ACA uses all observations where the variable of interest (`income`) is
observed, regardless of missingness on other variables.

```{r part-c-aca}
income_observed <- acs.mis$income[!is.na(acs.mis$income)]

mean_income_aca <- mean(income_observed)
se_income_aca   <- sd(income_observed) / sqrt(length(income_observed))

cat("Available-Case Analysis (ACA):\n")
cat("  Mean income :", round(mean_income_aca, 2), "\n")
cat("  SE          :", round(se_income_aca, 2), "\n")
cat("  n used      :", length(income_observed), "\n")
```

### Comparison

```{r part-c-comparison}
comparison <- data.frame(
  Method    = c("Full data (acs)", "CCA", "ACA"),
  Mean      = round(c(mean_income_full, mean_income_cca, mean_income_aca), 2),
  SE        = round(c(se_income_full, se_income_cca, se_income_aca), 2),
  n         = c(nrow(acs), nrow(acs_complete), length(income_observed))
)
comparison
```

**Discussion:**

- **CCA** uses fewer observations because it drops rows with missingness on
  *any* variable (both `income` and `married` contribute to exclusion). This
  leads to a larger standard error due to reduced sample size.
- **ACA** retains more observations (all rows where `income` is observed),
  providing a smaller standard error than CCA.
- If the mean from CCA and ACA differ substantially from the full-data mean,
  this suggests that the nonresponse mechanism is **not MCAR** (Missing
  Completely At Random). If the means are similar, it is consistent with MCAR
  — though it does not prove it.

---

# Part (d): Linear Regression of log(income)

**Question:** Run a linear regression of log(income) on all other variables
except `married`. Run this regression on both datasets, `acs` and `acs.mis`.
Discuss your findings. What does R do if you run the model on `acs.mis`? What
do you conclude regarding the nonresponse mechanism?

## Answer

### Regression on full data (`acs`)

```{r part-d-full}
# We need to handle income == 0 for log transformation.
# Common approach: use log(income + 1) or filter to income > 0.
# Here we use log(income) on observations with income > 0.

acs_pos <- acs[acs$income > 0, ]

model_full <- lm(log(income) ~ employment + race + age + gender + edu,
                 data = acs_pos)
summary(model_full)
```

### Regression on `acs.mis`

```{r part-d-mis}
acs_mis_pos <- acs.mis[!is.na(acs.mis$income) & acs.mis$income > 0, ]

model_mis <- lm(log(income) ~ employment + race + age + gender + edu,
                data = acs_mis_pos)
summary(model_mis)
```

### Comparing coefficients

```{r part-d-compare}
coef_comparison <- data.frame(
  Full_Data  = round(coef(model_full), 4),
  Missing_Data = round(coef(model_mis), 4),
  Difference = round(coef(model_full) - coef(model_mis), 4)
)
coef_comparison
```

**Discussion:**

- When running `lm()` on `acs.mis`, R **automatically performs complete-case
  analysis**: it silently drops all rows with any `NA` in the variables used by
  the model. Since `married` is excluded from the formula, the only source of
  NAs is `income` itself. R reports the number of deleted observations.
- If the regression coefficients from the two models are similar, this is
  consistent with the nonresponse in `income` being (at most) **MAR** —
  meaning that once we condition on the predictors in the model, the
  probability of `income` being missing does not depend on income itself.
- If the coefficients differ substantially, this could indicate that the
  nonresponse mechanism is **MNAR** (the probability of missingness depends on
  the unobserved income value).

---

# Part (e): Evaluating the Nonresponse Mechanism

**Question:** Since we have the fully observed data available, we can evaluate
the nonresponse mechanism by regressing the nonresponse indicator on the fully
observed data.

## Answer

### (e.a) Generate the missingness indicator

```{r part-e-indicator}
# R_income = 1 if income is missing in acs.mis, 0 otherwise
acs$R_income <- as.integer(is.na(acs.mis$income))

cat("Missingness indicator distribution:\n")
table(acs$R_income)
cat("\nProportion missing:", mean(acs$R_income), "\n")
```

### (e.b) Logistic regression of the missingness indicator

```{r part-e-logistic}
model_miss <- glm(R_income ~ income + employment + race + age + gender + edu + married,
                  data = acs, family = binomial)
summary(model_miss)
```

### Interpreting the results

```{r part-e-interpret}
# Odds ratios with confidence intervals
exp_coef <- exp(cbind(OR = coef(model_miss), confint(model_miss)))
round(exp_coef, 4)
```

**Discussion:**

- If **none** of the predictor variables (including `income` itself) are
  significant, this supports the hypothesis that the data are **MCAR** —
  missingness does not depend on any observed or unobserved quantity.
- If predictor variables **other than `income`** are significant (e.g., `age`,
  `employment`, `gender`) but `income` itself is not, this is consistent with
  **MAR** — missingness depends on observed covariates but not on the missing
  value itself.
- If `income` **is significant**, this provides evidence for **MNAR** —
  the probability of `income` being missing depends on the actual value of
  income. Note: in practice we would never have access to the true `income`
  for the missing cases, so this diagnostic is only possible here because we
  have the complete `acs` dataset.

> **Key caveat:** Even if `income` is significant in this logistic regression,
> we are testing MNAR *given* the other covariates. The practical distinction
> between MAR and MNAR always depends on which covariates are conditioned upon.

---

# Summary

| Part | Key Concept                          |
|------|--------------------------------------|
| (a)  | Identifying variables with NAs and computing nonresponse rates |
| (b)  | Baseline statistics from fully observed data |
| (c)  | CCA vs. ACA: impact of missing data on point estimates and SEs |
| (d)  | Regression under missingness; R's default behavior (`na.action`) |
| (e)  | Diagnosing the missingness mechanism using logistic regression |
