---
title: "Exercise Unit 4 – Analyzing Multiply Imputed Data"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../output"
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

The file `PASS.RData` contains the dataset `PASS.imp`, which consists of a subset of variables and records from the campus file of the PASS study. This exercise focuses on analyzing multiply imputed data using Rubin’s rules.

Variables included in the dataset:
- `hhsize`: Household size.
- `hhtype`: Household type.
- `hhincome`: Household income.
- `benefits`: Unemployment benefit recipient.
- `depind`: Deprivation index.
- `livspace`: Living space in square meters.

```{r load-data}
# Loading PASS.RData (contains PASS.imp) 
# and PASS_long.RData (contains PASS.long) as a fallback
load("../PASS.RData")
load("../PASS_long.RData")
```

---

# Part (a): Extracting Imputed Datasets

**Question:** Extract the multiply imputed datasets using the `complete` command from the `mice` package. Store the imputed datasets in the long format and call this dataset `PASS.long`.

## Answer

```{r part-a}
library(mice)

# Extracting the datasets in long format
# PASS.long <- complete(PASS.imp, action = "long")

# Since we already have PASS.long from the RData file, 
# we verify its structure:
str(PASS.long)
```

---

# Part (b): Computing Point Estimate (q-bar)

**Question:** Compute the final point estimate for the regression coefficient of "couple with children under 16" using `PASS.long`.

## Answer

We are interested in the regression:  
`log(depind) ~ log(hhincome) + hhtype + livspace + benefits`

We need to fit this model for each imputation $l = 1, \dots, m$ and average the coefficients.

```{r part-b}
# List of individual datasets based on .imp variable
m <- length(unique(PASS.long$.imp))
q_hat <- numeric(m)
u_hat <- numeric(m)

target_coef <- "hhtypecouple with children <16"

for (i in 1:m) {
  # Filter for the i-th imputation
  data_i <- PASS.long[PASS.long$.imp == i, ]
  
  # Fit the regression
  # Note: depind and hhincome are logged as requested
  mod <- lm(log(depind) ~ log(hhincome) + hhtype + livspace + benefits, data = data_i)
  
  # Extract point estimate and its variance (SE squared)
  sums <- summary(mod)
  q_hat[i] <- coef(mod)[target_coef]
  u_hat[i] <- sums$coefficients[target_coef, "Std. Error"]^2
}

# Final point estimate (average of individual estimates)
q_bar <- mean(q_hat)
q_bar
```

---

# Part (c): Computing Total Variance (T)

**Question:** Compute the estimated variance of the regression coefficient ($T$).

## Answer

Rubin's rule for total variance $T$ is:
$$T = \bar{U} + (1 + 1/m) B$$
where $\bar{U}$ is the within-imputation variance and $B$ is the between-imputation variance.

```{r part-c}
# Within-imputation variance (average of individual variances)
u_bar <- mean(u_hat)

# Between-imputation variance (variance of the point estimates)
b <- var(q_hat)

# Total variance
t_total <- u_bar + (1 + 1/m) * b
t_total
```

---

# Part (d): 95% Confidence Interval

**Question:** Compute a 95% confidence interval around your estimated regression coefficient.

## Answer

We need the degrees of freedom for the t-distribution. First define the relative increase in variance due to nonresponse:
$$r = \frac{(1 + 1/m) B}{\bar{U}}$$
Then the degrees of freedom are:
$$df = (m - 1) (1 + 1/r)^2$$

```{r part-d}
# Relative increase in variance
r <- (1 + 1/m) * b / u_bar

# Degrees of freedom (simple formula)
df <- (m - 1) * (1 + 1/r)^2

# Critical value (t-distribution)
alpha <- 0.05
t_crit <- qt(1 - alpha/2, df)

# Confidence interval
ci_low <- q_bar - t_crit * sqrt(t_total)
ci_high <- q_bar + t_crit * sqrt(t_total)

c(ci_low, ci_high)
```

---

# Part (e): P-value

**Question:** Can you compute the p-value for this coefficient?

## Answer

```{r part-e}
# T-statistic
t_stat <- q_bar / sqrt(t_total)

# Two-sided p-value
p_val <- 2 * (1 - pt(abs(t_stat), df))
p_val
```

---

# Part (f): Validation Using `pool()`

**Question:** Validate your results using the `pool` command from the `mice` package.

## Answer

```{r part-f}
# Using the full mids object with with() and pool()
fit <- with(PASS.imp, lm(log(depind) ~ log(hhincome) + hhtype + livspace + benefits))
pooled_results <- pool(fit)
summary(pooled_results)
```

**Discussion:**
The results from `pool()` should be very close to our manual calculations. Small differences might occur if `pool()` uses more advanced degrees of freedom adjustments (like Barnard-Rubin) or if there are differences in how the factor levels are handled.

---

# Part (g): Interpretation

**Question:** Interpret the results regarding the impact of having children under the age of 16 in the household.

## Answer

The coefficient for **`hhtypecouple with children <16`** represents the difference in the expected log(deprivation index) between a couple with children under 16 and the reference category (single households), holding income, living space, and benefits status constant.

If the coefficient is positive and significant (p < 0.05), it suggests that households with children under 16 experience higher deprivation scores compared to the reference group, after adjusting for other factors. Because the outcome is log-transformed, the percentage increase in the deprivation index can be approximated as $(\exp(\beta) - 1) \times 100\%$.
