---
title: "Exercise Unit 5 – Reproducibility of Multiple Imputation"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../output"
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This exercise illustrates the reproducibility argument in Multiple Imputation. We use the `PASS` data again, comparing results when generating $m=5$ vs. $m=50$ imputed datasets across 50 repeated calls to the `mice` command (using different random seeds).

The file `PASS_repeated_imp.RData` contains:
- `PASS.imp5`: A list of 50 `mids` objects, each with $m=5$ imputations.
- `PASS.imp50`: A list of 50 `mids` objects, each with $m=50$ imputations.

```{r load-data}
load("../PASS_repeated_imp.RData")
library(mice)
```

---

# Part (a): Repeated Analysis with $m=5$

**Question:** For each of the 50 `mice` outputs in `PASS.imp5`, run the linear regression:
`log(depind) ~ log(hhincome) + hhtype + livspace + benefits`.
Extract the coefficient for the "couple with children under 16" dummy, its variance (SE squared), and its p-value.

## Answer

```{r part-a}
# Define the target coefficient name
target_coef <- "hhtypecouple with children <16"

# Function to analyze a single mids object
analyze_mids <- function(imp_obj) {
  fit <- with(imp_obj, lm(log(depind) ~ log(hhincome) + hhtype + livspace + benefits))
  pooled <- pool(fit)
  res <- summary(pooled)
  
  # Extract row for target coefficient
  target_row <- res[res$term == target_coef, ]
  
  return(c(
    estimate = target_row$estimate,
    variance = target_row$std.error^2,
    p_value  = target_row$p.value
  ))
}

# Run analysis for all 50 iterations of m=5
results_m5 <- t(sapply(PASS.imp5, analyze_mids))
results_m5 <- as.data.frame(results_m5)

head(results_m5)
```

---

# Part (b): Repeated Analysis with $m=50$

**Question:** Compute the same results using `PASS.imp50`.

## Answer

```{r part-b}
# Run analysis for all 50 iterations of m=50
results_m50 <- t(sapply(PASS.imp50, analyze_mids))
results_m50 <- as.data.frame(results_m50)

head(results_m50)
```

---

# Part (c): Distribution of Regression Coefficients

**Question:** Compare the distribution of the estimated regression coefficients between `PASS.imp5` and `PASS.imp50`.

## Answer

```{r part-c}
summary_est <- data.frame(
  m5 = results_m5$estimate,
  m50 = results_m50$estimate
)
summary(summary_est)

# Graphical representation
boxplot(summary_est, main = "Distribution of Estimates (m=5 vs m=50)", 
        ylab = "Coefficient Estimate", col = c("lightblue", "lightgreen"))
```

**Discussion:**
The average estimate across all 50 iterations should be very similar for both $m=5$ and $m=50$. However, the coefficients for $m=50$ should show much less variation across the 50 iterations compared to $m=5$. This is because $m=50$ reduces the simulation error (Monte Carlo error) inherent in the imputation process.

---

# Part (d): Distribution of Estimated Variances

**Question:** Compare the distribution of the estimated MI variances between `PASS.imp5` and `PASS.imp50`.

## Answer

```{r part-d}
summary_var <- data.frame(
  m5 = results_m5$variance,
  m50 = results_m50$variance
)
summary(summary_var)

boxplot(summary_var, main = "Distribution of MI Variances (m=5 vs m=50)", 
        ylab = "Estimated Variance", col = c("lightblue", "lightgreen"))
```

**Discussion:**
Similar to the estimates, the MI variances estimated with $m=50$ should be much more stable (lower spread) across the 50 runs than those estimated with $m=5$.

---

# Part (e): Variance of the Estimates

**Question:** Compute the variance of the estimated regression coefficients across the 50 iterations for both datasets.

## Answer

```{r part-e}
var_of_estimates <- sapply(summary_est, var)
print(var_of_estimates)
```

**Discussion:**
The variance of the estimates across iterations is a direct measure of reproducibility. The variance for $m=5$ is expected to be significantly larger than for $m=50$, demonstrating why a higher $m$ is preferred for stable, reproducible results.

---

# Part (f): Distribution of P-values

**Question:** Compare the distribution of the estimated p-values.

## Answer

```{r part-f}
summary_p <- data.frame(
  m5 = results_m5$p_value,
  m50 = results_m50$p_value
)
summary(summary_p)

# Plotting the p-values
stripchart(summary_p, method = "jitter", vertical = TRUE, pch = 16, 
           col = c("blue", "red"), main = "P-value Distribution")
abline(h = 0.05, lty = 2, col = "darkgrey")
```

**Discussion:**
With $m=5$, you might find that some iterations yield p-values below 0.05 while others yield p-values above 0.05, purely due to the random noise of the imputation process. With $m=50$, the p-values are much more consistent, providing a more reliable basis for scientific conclusions.
